%% Master's thesis for Aalto University
%% Author: Viljami Aittomaki
%%
%% Written on official LaTeX template:
%% Original version by Luis Costa, changes by Perttu Puska


\documentclass[english,12pt,a4paper,pdftex,elec,utf8]{aaltothesis}


%%%% PAKCAGES %%%%

\usepackage{graphicx}
%% Use this if you write hard core mathematics, these are usually needed
\usepackage{amsmath,amsfonts,amssymb,amsbsy}

%% Use the macros in this package to change how the hyperref package below 
%% typesets its hypertext -- hyperlink colour, font, etc. See the package
%% documentation. It also defines the \url macro, so use the package when 
%% not using the hyperref package.
%\usepackage{url}

%% Use this if you want to get links and nice output. Works well with pdflatex.
\usepackage{hyperref}
\hypersetup{pdfpagemode=UseNone, pdfstartview=FitH,
  colorlinks=true,urlcolor=red,linkcolor=black,citecolor=black,
  pdftitle={MicroRNA regulation in breast cancer},pdfauthor={Viljami Aittom\"aki},
  pdfkeywords={Bayesian analysis, breast cancer, microRNA}}

%% User-added packages
\usepackage[numbers,square]{natbib}
\usepackage[parfill]{parskip}
\usepackage{longtable}
\usepackage[font={small,sf},labelfont={bf},format=plain,indention=.5cm]{caption}
\usepackage{subcaption}
\captionsetup[sub]{font={scriptsize,sf},labelfont={sf}}
\usepackage{pdflscape}
\setlength\LTcapwidth{\textwidth}
\setlength{\LTleft}{-20cm plus -1fill}
%\setlength{\LTleft}{-30cm}
\setlength{\LTright}{\LTleft}
\usepackage{colortbl}

%% Which sections to include
%\includeonly{introduction}







%%=========================================================
%% All that is printed on paper starts here
\begin{document}
\renewcommand{\thesissupervisorname}{Thesis supervisor}

%% Change the school field to specify your school if the automatically 
%% set name is wrong
% \university{aalto-yliopisto}
% \university{aalto University}
% \school{Sähkötekniikan korkeakoulu}
% \school{School of Electrical Engineering}

%% ONLY FOR M.Sc. AND LICENTIATE THESIS: Specify your department,
%% professorship and professorship code. 
%%
\department{Department of Computer Science}
\professorship{--}
%%

%% Valitse yksi näistä kolmesta
%%
%% Choose one of these:
%\univdegree{BSc}
\univdegree{MSc}
%\univdegree{Lic}

%% Your own name (should be self explanatory...)
\author{Viljami Aittom\"aki}

%% Your thesis title comes here and again before a possible abstract in
%% Finnish or Swedish . If the title is very long and latex does an
%% unsatisfactory job of breaking the lines, you will have to force a
%% linebreak with the \\ control character. 
%% Do not hyphenate titles.
%% 
\thesistitle{MicroRNA regulation in breast cancer -- \\a Bayesian analysis of expression data}

\place{Espoo,}
\date{7 October 2016}

%% Thesis supervisor 
\supervisor{Professor Aki Vehtari}

%% Thesis advisors(s).
\advisor{Senior Researcher Rainer Lehtonen}

%% Aalto logo: syntax:
%% \uselogo{aaltoRed|aaltoBlue|aaltoYellow|aaltoGray|aaltoGrayScale}{?|!|''}
%% Logo language is set to be the same as the document language.
%%
\uselogo{aaltoRed}{?}


%% Create the coverpage
\makecoverpage














%%=========================================================
%% English abstract.
%% All the information required in the abstract (your name, thesis title, etc.)
%% is used as specified above.
%% Specify keywords
%%
\keywords{Bayesian analysis, breast cancer, gene expression, microarray, microRNA, target prediction}
%% Abstract text
\begin{abstractpage}[english]

MicroRNAs are a class of small, non-coding RNAs, which regulate gene
expression post-transcriptionally. They downregulate genes by targeting
messenger RNA transcripts and causing their degradation and inhibition of
translation. Research has revealed microRNAs to participate in diverse
cellular functions, such as differentiation and apoptosis, and many
pathological processes, including cancer. \\

Identification of microRNA target genes is crucial in understanding their
function in cell biology and disease. A wide range of methods have been
proposed for computational prediction of microRNA targets. Early target
prediction methods used sequence information, while recent tools have
integrated expression measurements of target genes and microRNAs. A limited
number of studies have integrated protein, gene and microRNA
expression for target prediction. \\

Breast cancer is the most common cancer in women and a significant cause of
morbidity and mortality globally. Analyses of gene expression data have
provided insight into the pathogenesis of breast cancer, and intrinsic
subtypes correlating with prognosis have been identified. A range of microRNAs
have been indicated to contribute to breast cancer pathogenesis. \\

% provides a review of microRNA biology and target prediction.
In this thesis, a
recent Bayesian variable selection method was applied for uncovering putative
microRNA targets in breast cancer. The proposed model integrated
protein, gene and microRNA expression data. Results
were compared with another popular prediction method.
%and a recently published analysis of the same data.
Analyses showed that the proposed method is applicable to microRNA target
prediction. Limitations and refinements of the method and study are discussed,
and the importance of an integrative approach is highlighted.

\end{abstractpage}

%% Force a new page so that the Finnish abstract starts on a new page
\newpage

%% Abstract in Finnish.  Delete if you don't need it. 
\thesistitle{MikroRNA-säätely rintasyövässä -- ekspressiodatan bayesilainen analyysi}
\supervisor{Professori Aki Vehtari}
\advisor{Dosentti Rainer Lehtonen}
\degreeprogram{Electronics and electrical engineering}
\department{Tietotekniikan laitos}
\professorship{--}
%% Avainsanat
\keywords{bayesilainen analyysi, geeniekspressio, kohdegeenien ennustaminen, mikroRNA, mikrosiru, rintasy\"op\"a}
%% Tiivistelman tekstiosa
\begin{abstractpage}[finnish]

MikroRNA:t ovat lyhyitä RNA-molekyylejä, jotka säätelevät geeniekspressiota
sitoutumalla lähetti-RNA-molekyyleihin estäen siten niiden translaation
proteiiniksi. Aiemmat tutkimukset ovat osoittaneet, että mikroRNA:t
osallistuvat monipuolisesti solujen toiminnan säätelyyn, kuten erilaistumiseen ja
apoptoosiin, ja ovat osallisena monien tautien, kuten syövän synnyssä. \\

MikroRNA:n säätelemien kohdegeenien tunnistaminen on olennainen
askel mikroRNA:n toiminnan ymmärtämisessä. Kohdegeenien
ennustamiseen on kehitetty lukuisia laskennallisia menetelmiä. Varhaiset
menetelmät perustuivat RNA-sekvenssien vertailuun. Uudemmat työkalut
yhdistävät geeni- ja mikroRNA-ekspressiodataa kohdegeenien tunnistamiseksi.
Proteiini-, geeni- ja mikroRNA-ekspressiota yhdistäviä
kohdegeenien tunnistamiseen tähtääviä tutkimuksia on julkaistu toistaiseksi
suhteellisen vähän. \\

Rintasyöpä on naisten yleisin syöpä ja merkittävä sairastavuuden ja
kuolleisuuden aiheuttaja maailmanlaajuisesti. Geeniekspressiodatan analysointi
on lisännyt tietoa rintasyövän synnystä, ja geeniekspressioon perustuen on
kyetty tunnistamaan rintasyövän alatyyppejä, jotka korreloivat syövän
ennusteeseen. MikroRNA:n on todettu olevan osatekijä rintasyövän synnyssä. \\

Tässä diplomityössä sovellettiin äskettäin julkaistua bayesilaista
muuttujavalintamenetelmää mikroRNA-molekyylien kohdegeenien ennustamiseen
rintasyövässä. Tähän tarkoitukseen käytettiin proteiini-, geeni- ja mikroRNA-
ekspressiodataa. Tulokset osoittivat, että menetelmä soveltuu kohdegeenien
ennustamiseen. Työssä esitetään vaihtoehtoja mallin jatkokehittämiseksi.

\end{abstractpage}














%%=========================================================
%% Preface

%% Force new page so that the preface starts from a new page
\newpage

\mysection{Preface}

This Master's Thesis and the related research were done in the Systems Biology
of Drug Resistance in Cancer research group within the Genome-Scale Biology
research program of the University of Helsinki. I am grateful to Professor
Sampsa Hautaniemi for suggesting this work and providing facilities and also
for several years of interesting work in his group.

I want to thank Professor Aki Vehtari for supervision and suggesting the
method proposed in this work as well as for advice in implementing it. I would
like to thank my advisors Professor Sampsa Hautaniemi and Senior Researcher
Rainer Lehtonen for their guidance and trust and the time they have invested
into this work. I want to thank Juho Piironen for help with the modeling. I
would also like to thank Professor Antti Vaheri for discussions.

I owe a lifetime of gratitude to my parents for their endless love and
encouragement. Even in the darkest of moments, they have never stopped
believing in me. I also want to thank my mother for discussions relating
to genetics.

Finally, I wish to send a thank you and all of my love to all my family and
friends for their support during the writing of this thesis and in other
endeavors in my life. Special thanks to my friend Heidi for pushing me forward
and sharing cookies. \\

\vspace{5cm}
Espoo, 7 October 2016

\vspace{5mm}
{\hfill Viljami Aittom\"aki \hspace{1cm}}


%% Force new page after preface
\newpage


%% Table of contents. 
\thesistableofcontents













%%=========================================================
%% Symbols and abbreviations
\mysection{Symbols and abbreviations}

\subsection*{Symbols}

\begin{tabular}{ll}
$\beta$                    & Regression coefficients for explanatory variables \\
$\left\| \beta \right\|_1$ & The 1-norm of $\beta$ \\
$\textup{E}_{\theta}(x)$   & Expectation of random variable $x$ over parameters $\theta$ \\
$n$                        & Number of observations \\
$\textup{N}(\mu,\sigma^2)$ & Multivariate normal distribution with mean $\mu$ and variance $\sigma^2$ \\
$p_n$                      & Number of (assumed) true explanatory variables in variable seĺection \\
$p(y)$                     & Probability density of $x$ \\
$p(y,\theta)$              & Joint probability of $y$ and $\theta$ \\
$p(y|\theta)$              & Conditional probability of $y$ given $\theta$ \\
$\theta$                   & Parameters of a probability model \\
$x$                        & Explanatory variable (or microRNA expression vector) \\
$X$                        & Matrix of explanatory variables (or microRNA expression vectors) \\
$y$                        & Outcome variable (or protein expression vector) \\
$y \sim \textup{N}(.)$     & Random variable $y$ has probability distribution $N(.)$ \\
$y \propto x$              & $y$ is proportional to $x$, up to a constant factor \\
$z$                        & mRNA expression vector
\end{tabular}

% \subsection*{Expressions}
% \begin{tabular}{ll}
% $\nabla \times \mathbf{A}$                 & curl of vectorin $\mathbf{A}$\\
% $\displaystyle\frac{\mbox{d}}{\mbox{d} t}$ & derivative with respect to variable $t$\\[3mm]
% $\displaystyle\frac{\partial}{\partial t}$ & partial derivative with respect to variable $t$ \\[3mm]
% $\sum_i $                      & sum over index $i$\\
% $\mathbf{A} \cdot \mathbf{B}$  & dot product of vectors $\mathbf{A}$ and $\mathbf{B}$
% \end{tabular}

\subsection*{Abbreviations}

\begin{tabular}{lp{11cm}}
bp          & Base pairs (as a measure of double-stranded sequence length) \\
HMM         & Hidden Markov model \\
lasso       & Least absolute shrinkage and selection operator \\
LPD         & log predictive density \\
MCMC        & Markov chain Monte Carlo \\
miRNA       & MicroRNA \\
MLPD        & Mean log predictive density \\
MLR         & Multivariate linear regression \\
mRNA        & Messenger RNA \\
NGS         & Next-generation sequencing \\
nt          & Nucleotides (as a measure of sequence length) \\
PCR         & Polymerase chain reaction \\
PPVS        & Projection predictive variable selection \\
qPCR        & Quantitative PCR \\
RISC        & RNA-induced silencing complex \\
RNAi        & RNA interference \\
RPPA        & Reverse-phase protein array \\
SVM         & Support vector machine \\
UTR         & Untranslated region at the beginning (5') or end (3') of a messenger RNA \\
WHO         & World Health Organization \\
\end{tabular}


%% Tweaks the page numbering to meet the requirement of the thesis format:
%% Begin the pagenumbering in Arabian numerals (and leave the first page
%% of the text body empty, see \thispagestyle{empty} below).
%% Additionally, force the actual text to begin on a new page with the 
%% \clearpage command.
%% \clearpage is similar to \newpage, but it also flushes the floats (figures
%% and tables).
%% There is no need to change these
%%
\cleardoublepage
\storeinipagenumber
\pagenumbering{arabic}
\setcounter{page}{1}








%%=========================================================
%% Text body begins

\include{introduction}

%\clearpage (\include implicity does a \clearpage)
\include{gene-expression}

\include{micrornas}

\include{cancer}

\include{microrna-target-identification}

\include{bayesian-analysis}

\include{material-methods}

\include{results}

\include{discussion}




%%=========================================================
%% References

\clearpage
%% The \phantomsection command is nessesary for hyperref to jump to the 
%% correct page, in other words it puts a hyper marker on the page.
\phantomsection

\addcontentsline{toc}{section}{\refname}
%\addcontentsline{toc}{section}{References}

\bibliographystyle{apalike}
\bibliography{dippa}




%%=========================================================
%% Appendices

\clearpage
\thesisappendix

\include{appendix-model-table}

\include{appendix-model-sizes}

\include{appendix-QC-plots}




%%=========================================================
%% The End  ==DDDDDDD

\end{document}
